Explanation-based learning of action models

### 2019-03-06 doc/lecture-notes-EBL/ebl.tex
updated sections 1,2,3,4 and 6 lecture notes on ebl

%%% Defintions For this paper
\usepackage[numbers]{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{arrows,automata}
\usepackage{comment}

@@

Considering only the syntactic constraints, the size of the space of possible \strips\ models is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$ because one element in $\mathcal{I}_{\xi,\Psi}$ can appear both in the preconditions and effects of $\xi$. The belonging of an $e \in \mathcal{I}_{\Psi,\xi}$ to the preconditions, positive effects or negative effects of $\xi$ is handled with a refined propositional encoding that uses fluents of two types, $pre\_\xi\_e$ and $eff\_\xi\_e$, instead of the three fluents used in the original compilation~\cite{aineto2018learning}. The four possible combinations of these two fluents are sumarized in Figure \ref{fig:combinations}. This compact encoding allows for a more effective exploitation of the syntactic constraints, and also yields the solution space of $\Lambda$ to be the same as its search space.

\begin{figure}
	\begin{footnotesize}
		\begin{tabular}{c@{\hskip .3in}  c}
	{\bf Encoding} & {\bf Meaning}\\\hline
$\neg pre\_\xi\_e \wedge \neg eff\_\xi\_e $& $e$ belongs neither to the preconditions nor effects of $\xi$ \\
             & ($e \notin pre(\xi) \wedge e \notin add(\xi) \wedge e \notin del(\xi)$)\\\\
$pre\_\xi\_e \wedge \neg eff\_\xi\_e $& $e$ is only a precondition of $\xi$\\
               &  ($e \in pre(\xi) \wedge e \notin add(\xi) \wedge e \notin del(\xi)$) \\\\
$\neg pre\_\xi\_e \wedge eff\_\xi\_e $& $e$ is a positive effect of $\xi$ \\
               &  ($e \notin pre(\xi) \wedge e \in add(\xi) \wedge e \notin del(\xi)$) \\\\
$pre\_\xi\_e \wedge eff\_\xi\_e  $& $e$ is a negative effect of $\xi$ \\
               &  ($e \in pre(\xi) \wedge e \notin add(\xi) \wedge e \in add(\xi)$)
		\end{tabular}
	\end{footnotesize}
	\caption{\small Combinations of the fluent propositional encoding and their meaning}
	\label{fig:combinations}
\end{figure}

@@

\begin{enumerate}
\item {\bf To build the action models of $\mathcal{M}'$}. $\pi_\Lambda$ comprises a first block of actions (plan {\em prefix}) that set the predicates $p\in \Psi_{\xi}$ of $pre(\xi)$, $del(\xi)$ and $add(\xi)$ for each $\xi\in\mathcal{M}$.
\item {\bf To validate the action models of $\mathcal{M}'$}. $\pi_\Lambda$ also comprises a second block of actions (plan {\em postfix}) which is aimed at validating the observation ${\mathcal O}$ with the built action models $\mathcal{M}'$.
\end{enumerate}

Given a learning task $\Lambda=\tup{\mathcal{M},{\mathcal O}}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$ such that:
\begin{itemize}
\item $F_{\Lambda}$ extends $F$ with the model fluents to represent the preconditions and effects of each $\xi\in\mathcal{M}$ as well as some other fluents to keep track of the validation of ${\mathcal O}$. Specifically, $F_{\Lambda}$ contains:
\begin{itemize}
\item The set of fluents obtained from $s_0^o$; i.e., $F$.
\item The model fluents $pre_p(\xi)$ and $eff_p(\xi)$, for every $\xi \in \mathcal{M}$ and $p\in \Psi_{\xi}$, defined as explained in section~\ref{sec:strips-space}.
\item A set of fluents $\{test_j\}_{0\leq j\leq m}$, to point at the state observation $s_j^o\in {\mathcal O}$ where the action model is validated. In the example of Figure~\ref{fig:observation} two tests are required to validate the programmed action model, one corresponding to the initial state and the second one corresponding to the final state.
\item A fluent, $mode_{prog}$, to indicate whether action models are being programmed or validated.
\end{itemize}

@@

\item $G_{\Lambda}$ includes the positive literal $test_m$. When this goal is achieved by the solution plan $\pi_\Lambda$, we will be certain that the action models of $\mathcal{M'}$ are validated in the input observation.

\item $A_{\Lambda}$ includes three types of actions that give rise to the actions of $\pi_\Lambda$.
\begin{enumerate}
\item Actions for {\em inserting} a component (precondition or effect) in $\xi \in \mathcal{M}$ following the syntactic constraints of \strips\ models. These actions will form the prefix of the solution plan $\pi_\Lambda$. Among the \emph{inserting} actions, we find:
\begin{itemize}
\item Actions which support the addition of a {\em precondition} $p\in \Psi_{\xi}$ to the action model $\xi\in\mathcal{M}$. A precondition $p$ is inserted in $\xi$ when neither $pre_p$ nor $eff_p$ exist in $\xi$.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertPre_{p,\xi}})=&\{\neg pre_{p}(\xi), \neg eff_{p}(\xi), mode_{prog}\},\\
\cond(\mathsf{insertPre_{p,\xi}})=&\{\emptyset\}\rhd\{pre_{p}(\xi)\}.
\end{align*}
\end{small}

\item Actions which support the addition of a {\em negative} or {\em positive} effect $p\in \Psi_{\xi}$ to the action model $\xi\in\mathcal{M}$. A positive effect is inserted in $\xi$ under the same conditions of a precondition insertion, and a negative effect is inserted in $\xi$ when neither $del_p$ nor $add_p$ appear in $\xi$ but $pre_p$ does.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertEff_{p,\xi}})=&\{\neg del_{p}(\xi),\neg add_{p}(\xi), mode_{prog}\},\\
\cond(\mathsf{insertEff_{p,\xi}})=&\{pre_{p}(\xi)\}\rhd\{del_{p}(\xi)\},\\
& \{\neg pre_{p}(\xi)\}\rhd\{add_{p}(\xi)\}.
\end{align*}
\end{small}
\end{itemize}

For instance, given $name(\xi)=${\tt{\small stack}} and $C_{pre-stack}=\{${\tt{\small (pre\_stack\_holding\_v1),(pre\_stack\_holding\_v2), (pre\_stack\_on\_v1\_v2),(pre\_stack\_clear\_v1),(pre\_stack\_clear\_v1),}}$\ldots \}$, the insertion of each item $c \in C_{pre-stack}$ in $\xi$ will generate a different alternative in the search space when solving $P_{\Lambda}$ as long as $c \notin pre(\xi)$,  $c \notin add(\xi)$ and $c \notin del(\xi)$. The same applies to effects with respect to sets $C_{add-stack}$ and $C_{del-stack}$ that would include all fluents starting with prefix  {\tt{\small add}} and {\tt{\small del}}, respectively.

@@

\item Actions for {\em applying} the action models $\xi\in\mathcal{M}$ built by the insert actions and bounded to objects $\omega\subseteq\Omega^{ar(\xi)}$. Since action headers are known, the variables $pars(\xi)$ are bounded to the objects in $\omega$ that appear in the same position.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{p}(\xi)\implies p(\omega)\}_{\forall p\in\Psi_\xi},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{del_{p}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi_\xi},\\
&\{add_{p}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi\xi},\\
&\{mode_{prog}\}\rhd\{\neg mode_{prog}\}.
\end{align*}
\end{small}

    These actions will be part of the postfix of the plan $\pi_\Lambda$ and they determine the application of the learned action models according to the values of the model fluents in the current state configuration}. Figure~\ref{fig:compilation} shows the PDDL encoding of {\tt{\small (apply\_stack)}} for applying the action model of the {\em stack} operator. \textcolor[rgb]{1.00,0.00,0.00}{Let's assume the action {\tt{\small (apply\_stack blockB blockA)}} is in $\pi_\Lambda$. Executing this action in a state $s$ implies activating the preconditions and effects of {\tt{\small (apply\_stack)}} according to the values of the model fluents in $s$. For example, if  $\{${\tt{\small (pre\_stack\_holding\_v1),(pre\_stack\_clear\_v2)}}$\} \subset s$ then it must be checked that positive literals {\tt{\small (holding blockB)}} and {\tt{\small (clear blockA)}} hold in $s$. Otherwise, a different set of precondition literals will be checked. The same applies to the conditional effects, generating the corresponding literals according to the values of the model fluents of $s$.

@@

\begin{figure}[hbt!]
\begin{center}
\begin{scriptsize}
\begin{verbatim}
(:action apply_stack
  :parameters (?o1 - object ?o2 - object)
  :precondition
   (and (or (not (pre_stack_on_v1_v1)) (on ?o1 ?o1))
        (or (not (pre_stack_on_v1_v2)) (on ?o1 ?o2))
        (or (not (pre_stack_on_v2_v1)) (on ?o2 ?o1))
        (or (not (pre_stack_on_v2_v2)) (on ?o2 ?o2))
        (or (not (pre_stack_ontable_v1)) (ontable ?o1))
        (or (not (pre_stack_ontable_v2)) (ontable ?o2))
        (or (not (pre_stack_clear_v1)) (clear ?o1))
        (or (not (pre_stack_clear_v2)) (clear ?o2))
        (or (not (pre_stack_holding_v1)) (holding ?o1))
        (or (not (pre_stack_holding_v2)) (holding ?o2))
        (or (not (pre_stack_handempty)) (handempty)))
  :effect
   (and (when (del_stack_on_v1_v1) (not (on ?o1 ?o1)))
        (when (del_stack_on_v1_v2) (not (on ?o1 ?o2)))
        (when (del_stack_on_v2_v1) (not (on ?o2 ?o1)))
        (when (del_stack_on_v2_v2) (not (on ?o2 ?o2)))
        (when (del_stack_ontable_v1) (not (ontable ?o1)))
        (when (del_stack_ontable_v2) (not (ontable ?o2)))
        (when (del_stack_clear_v1) (not (clear ?o1)))
        (when (del_stack_clear_v2) (not (clear ?o2)))
        (when (del_stack_holding_v1) (not (holding ?o1)))
        (when (del_stack_holding_v2) (not (holding ?o2)))
        (when (del_stack_handempty) (not (handempty)))
        (when (add_stack_on_v1_v1) (on ?o1 ?o1))
        (when (add_stack_on_v1_v2) (on ?o1 ?o2))
        (when (add_stack_on_v2_v1) (on ?o2 ?o1))
        (when (add_stack_on_v2_v2) (on ?o2 ?o2))
        (when (add_stack_ontable_v1) (ontable ?o1))
        (when (add_stack_ontable_v2) (ontable ?o2))
        (when (add_stack_clear_v1) (clear ?o1))
        (when (add_stack_clear_v2) (clear ?o2))
        (when (add_stack_holding_v1) (holding ?o1))
        (when (add_stack_holding_v2) (holding ?o2))
        (when (add_stack_handempty) (handempty))
        (when (modeProg) (not (modeProg)))))
\end{verbatim}
\end{scriptsize}
 \caption{\small PDDL action for applying an already programmed model for $stack$ (implications are coded as disjunctions).}
\label{fig:compilation}
\end{center}
\end{figure}

@@

\item Actions for {\em validating} the partially observed state $s_j\in\tau$, {\tt\small $1\leq j< m$}. These actions are also part of the postfix of the solution plan $\pi_\Lambda$ and they are aimed at checking that the observable data of the input plan trace $\tau$ follows after the execution of the apply actions.

@@

There will be a validate action in $\pi_\Lambda$ for every observed state in $\tau$. The position of the validate actions in $\pi_\Lambda$ will be determined by the planner by checking that the state resulting after the execution of an apply action comprises the observed state $s_j\in\tau$.

@@

In some contexts, it is reasonable to assume that some parts of the action model are known and so there is no need to learn the entire model from scratch \cite{ZhuoNK13}. In {\sc FAMA}, when an action model $\xi$ is partially specified, the known preconditions and effects are encoded as fluents $pre_p(\xi)$, $del_p(\xi)$ and $add_p(\xi)$ set to true in the initial state $I_{\Lambda}$. In this case, the corresponding insert actions, $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertEff_{p,\xi}}$, become unnecessary and are removed from $A_{\Lambda}$, thereby making the classical planning task $P_{\Lambda}$ easier to be solved.

So far we have explained the compilation for learning from a single input trace. However, the compilation is extensible to the more general case $\Lambda=\tup{\mathcal{M},\mathcal{T}}$, where $\mathcal{T}=\{\tau_1,\ldots,\tau_k\}$ is a set of plan traces. Taking this into account, a small modification is required in our compilation approach. In particular, the actions in $P_{\Lambda}$ for {\em validating} the last state $s_m^t\in \tau_t$, {\tt\small $1\leq t\leq k$} of a plan trace $\tau_t$ reset the current state and the current plan. These actions are now redefined as:

\begin{small}
	\begin{align*}
	\hspace*{7pt}\pre(\mathsf{validate_{j}})=&s_m^t\cup\{test_{j-1}\}\cup \{\neg mode_{prog}\},\\
	\cond(\mathsf{validate_{j}})=&\{\emptyset\}\rhd\{\neg test_{j-1},test_j\} \cup \\
	&\{\neg f\}_{\forall f\in s_m^t, f \notin s_0^{t+1}}\cup \{f\}_{\forall f\in s_0^{t+1}, f \notin s_m^t},\\
	&\{\neg f\}_{\forall f\in F_{\pi_t}}\cup \{f\}_{\forall f\in F_{\pi_{t+1}}}.\\
	\end{align*}
\end{small}

Finally, we will detail the composition of a solution plan $\pi_\Lambda$ to a planning task $P_\Lambda$ and the mechanism to extract the action models of $\mathcal{M}'$ from $\pi_\Lambda$. The plan of Figure~\ref{fig:plan-lplan} shows a solution to the task $P_{\Lambda}$ that encodes a learning task $\Lambda=\tup{\mathcal{M},\tau}$ for obtaining the action models of the {\em blocksworld} domain, where the models for {\tt\small pickup}, {\tt\small putdown} and {\tt\small unstack} are already specified in $\mathcal{M}$. Therefore, the plan shows the insert actions and validate action for the action model {\tt\small stack} using the input plan trace of Figure~\ref{fig:example-plans}. Plan steps $00-01$ insert the preconditions of the {\tt\small stack} model, steps $02-06$ insert the action model effects, and steps $07-11$ form the plan postfix that applies the action models (only the {\tt\small stack} model is learned) and validates the result in the plan trace of Figure~\ref{fig:example-plans}.

\begin{figure}[hbt!]
	{\footnotesize\tt
		{\bf 00} : (insert\_pre\_stack\_holding\_v1) \\
		01 : (insert\_pre\_stack\_clear\_v2)\\
		{\bf 02} : (insert\_eff\_stack\_clear\_v1)\\
		03 : (insert\_eff\_stack\_clear\_v2)\\
		04 : (insert\_eff\_stack\_handempty)\\
		05 : (insert\_eff\_stack\_holding\_v1)\\
		06 : (insert\_eff\_stack\_on\_v1\_v2)\\
		{\bf 07} : (apply\_unstack blockB blockA i1 i2)\\
		08 : (apply\_putdown blockB i2 i3)\\
		09 : (apply\_pickup blockA i3 i4)\\
		10 : (apply\_stack blockA blockB i4 i5)\\
		{\bf 11} : (validate\_1)
	}
	\caption{\small Plan for programming and validating the $stack$ action model (using the plan trace $\tau$ of Figure~\ref{fig:example-plans}) as well as previously specified action models for $pickup$, $putdown$ and $unstack$.}
	\label{fig:plan-lplan}
\end{figure}

Given a solution plan $\pi_\Lambda$ that solves $P_{\Lambda}$, the set of action models $\mathcal{M}'$ that solves $\Lambda=\tup{\mathcal{M},\tau}$ are computed in linear time and space. In order to do so, $\pi_\Lambda$ is executed in the initial state $I_{\Lambda}$ and the action model $\mathcal{M}'$ will be given by the fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$ that are set to true in the last state reached by $\pi_\Lambda$, $s_g=\theta(I_\Lambda,\pi_\Lambda)$. For each $\xi \in \mathcal{M'}$, we build the sets of preconditions, positive effects and negative effects as follows:

>>>>

%%% Defintions For this paper
\usepackage[numbers]{natbib}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{arrows,automata}
\usepackage{comment}

@@

Considering only the syntactic constraints, the size of the space of possible \strips\ models is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$ because one element in $\mathcal{I}_{\xi,\Psi}$ can appear both in the preconditions and effects of $\xi$. The belonging of a $p \in \mathcal{I}_{\Psi,\xi}$ to the preconditions, positive effects or negative effects of $\xi$ is handled with a refined propositional encoding that uses fluents of two types, $pre_{p,\xi}$ and $eff_{p,\xi}$, instead of the three fluents used in the original compilation~\cite{aineto2018learning}. The four possible combinations of these two fluents are sumarized in Figure \ref{fig:combinations}. This compact encoding allows for a more effective exploitation of the syntactic constraints, and also yields the solution space of $\Lambda$ to be the same as its search space.

\begin{figure}
	\begin{footnotesize}
		\begin{tabular}{c@{\hskip .2in} |@{\hskip .1in} c}
	{\bf Encoding} & {\bf Meaning}\\\hline
$\neg pre_{p,\xi} \wedge \neg eff_{p,\xi} $& $p$ belongs neither to the preconditions nor effects of $\xi$ \\
             & ($p \notin pre(\xi) \wedge p \notin add(\xi) \wedge p \notin del(\xi)$)\\\\
$pre_{p,\xi} \wedge \neg eff_{p,\xi} $& $p$ is only a precondition of $\xi$\\
               &  ($p \in pre(\xi) \wedge p \notin add(\xi) \wedge p \notin del(\xi)$) \\\\
$\neg pre_{p,\xi} \wedge eff_{p,\xi} $& $p$ is a positive effect of $\xi$ \\
               &  ($p \notin pre(\xi) \wedge p \in add(\xi) \wedge p \notin del(\xi)$) \\\\
$pre_{p,\xi} \wedge eff_{p,\xi} $& $p$ is a negative effect of $\xi$ \\
               &  ($p \in pre(\xi) \wedge p \notin add(\xi) \wedge p \in add(\xi)$)
		\end{tabular}
	\end{footnotesize}
	\caption{\small Combinations of the fluent propositional encoding and their meaning}
	\label{fig:combinations}
\end{figure}

To illustrate better this encoding, Figure~\ref{fig:propositional} shows the PDDL encoding of the {\tt\small stack(?v1,?v2)} schema and our propositional representation for this same schema with {\tt\small pre\_e\_stack} and {\tt\small eff\_e\_stack} fluents ($e\in{\mathcal I}_{\Psi,stack}$).

\begin{figure}
  \begin{small}  
  \begin{verbatim}
(:action stack
   :parameters (?v1 ?v2)
   :precondition (and (holding ?v1) (clear ?v2))
   :effect (and (not (holding ?v1)) (not (clear ?v2)) (clear ?v1) (handempty) (on ?v1 ?v2)))

(pre_holding_v1_stack) (pre_clear_v2_stack)
(eff_holding_v1_stack) (eff_clear_v2_stack)
(eff_clear_v1_stack) (eff_handempty_stack) (eff_on_v1_v2_stack)
  \end{verbatim}           
  \end{small}  
 \caption{\small PDDL encoding of the {\tt\small stack(?v1,?v2)} schema and our propositional representation for this same schema.}
\label{fig:propositional}
\end{figure}

@@

\begin{enumerate}
\item {\bf To build the action models of $\mathcal{M}'$}. $\pi_\Lambda$ comprises a first block of actions (plan {\em prefix}) that set the predicates $p\in \Psi_{\xi}$ of $pre(\xi)$, $del(\xi)$ and $add(\xi)$ for each $\xi\in\mathcal{M}$.
\item {\bf To validate the action models of $\mathcal{M}'$}. $\pi_\Lambda$ also comprises a second block of actions (plan {\em postfix}) which is aimed at {\em explaining} the observation ${\mathcal O}$ with the built action models $\mathcal{M}'$.
\end{enumerate}

Given a learning task $\Lambda=\tup{\mathcal{M},{\mathcal O}}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$ such that:
\begin{itemize}
\item $F_{\Lambda}$ extends $F$ with the model fluents to represent the preconditions and effects of each $\xi\in\mathcal{M}$ as well as some other fluents to keep track of the validation of ${\mathcal O}$. Specifically, $F_{\Lambda}$ contains:
\begin{itemize}
\item The set of fluents obtained from $s_0^o$; i.e., $F$.
\item The model fluents $pre_{p,\xi}$ and $eff_{p,\xi}$, for every $p\in \Psi_{\xi}$ and $\xi \in \mathcal{M}$, defined as explained in section~\ref{sec:strips-space}.
\item A set of fluents $\{test_j\}_{0\leq j\leq m}$, to point at the state observation $s_j^o\in {\mathcal O}$ where the action model is validated. In the example of Figure~\ref{fig:observation} two tests are required to validate the programmed action model, one corresponding to the initial state and the second one corresponding to the final state.  
\item A fluent, $mode_{prog}$, to indicate whether action models are being programmed or validated and a fluent {\small\tt invalid} to indicate that the programmed action model is inconsistent with the input observation.
\end{itemize}

@@

\item $G_{\Lambda}$ includes the positive literal $test_m$ and the negative literal $\neg${\small\tt invalid}. When these goals are achieved by the solution plan $\pi_\Lambda$, we will be certain that the action models of $\mathcal{M'}$ are validated in the input observation.

\item $A_{\Lambda}$ includes three types of actions that give rise to the actions of $\pi_\Lambda$.
\begin{enumerate}
\item Actions for {\em inserting} a component (precondition or effect) in $\xi \in \mathcal{M}$ following the syntactic constraints of \strips\ models. These actions will form the prefix of the solution plan $\pi_\Lambda$. Among the \emph{inserting} actions, we find:
\begin{itemize}
\item Actions which support the addition of a {\em precondition} $p\in \Psi_{\xi}$ to the action model $\xi\in\mathcal{M}$. A precondition $p$ is inserted in $\xi$ when neither $pre_p$ nor $eff_p$ exist in $\xi$.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertPre_{p,\xi}})=&\{\neg pre_{p,\xi}, \neg eff_{p,\xi}, mode_{prog}\},\\
\cond(\mathsf{insertPre_{p,\xi}})=&\{\emptyset\}\rhd\{pre_{p,\xi}\}.
\end{align*}
\end{small}

\item Actions which support the addition of a effect $p\in \Psi_{\xi}$ to the action model $\xi\in\mathcal{M}$. 
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{insertEff_{p,\xi}})=&\{\neg eff_{p,\xi}, mode_{prog}\},\\
\cond(\mathsf{insertEff_{p,\xi}})=&\{\emptyset\}\rhd\{eff_{p,\xi}\}
\end{align*}
\end{small}
\end{itemize}

For instance, given $name(\xi)=${\tt{\small stack}} and $F_{pre,stack}=\{${\tt{\small (pre\_stack\_holding\_v1),} {\tt\small (pre\_stack\_holding\_v2), (pre\_stack\_on\_v1\_v2),(pre\_stack\_clear\_v1)}, {\tt\small(pre\_stack\_clear\_v1)}}, $\ldots \}$, the insertion of each item $p \in F_{pre,stack}\subseteq F_{\Lambda}$ in $\xi$ will generate a different alternative in the search space when solving $P_{\Lambda}$. The same applies to effects $F_{eff,stack}=\{${\tt{\small (eff\_stack\_holding\_v1),(eff\_stack\_holding\_v2), (eff\_stack\_on\_v1\_v2),(eff\_stack\_clear\_v1),(eff\_stack\_clear\_v1),}}$\ldots \}$.

@@

\item Actions for {\em applying} the action models $\xi\in\mathcal{M}$ built by the insert actions and bounded to objects $\omega\subseteq\Omega^{ar(\xi)}$. These actions will be part of the postfix of the plan $\pi_\Lambda$ and they determine the application of the learned action models according to the values of the model fluents in the current state configuration. Since action headers are known, the variables $pars(\xi)$ are bounded to the objects in $\omega$ that appear in the same position.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{\},\\
\cond(\mathsf{apply_{\xi,\omega}})=& \{pre_{p,\xi}\wedge eff_{p,\xi}\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi_\xi},\\
& \{\neg pre_{p,\xi}\wedge eff_{p,\xi}\}\rhd\{p(\omega)\}_{\forall p\in\Psi\xi},\\
& \{pre_{p,\xi}\wedge \neg p(\omega)\}_{\forall p\in\Psi_\xi}\rhd\{invalid\},\\
&\{mode_{prog}\}\rhd\{\neg mode_{prog}\}.
\end{align*}
\end{small}

Figure~\ref{fig:compilation} shows the PDDL encoding of {\tt{\small (apply\_stack)}} for applying the action model of the {\em stack} operator. Let's assume the action {\tt{\small (apply\_stack blockB blockA)}} is in $\pi_\Lambda$. Executing this action in a state $s$ implies activating the preconditions and effects of {\tt{\small (apply\_stack)}} according to the values of the model fluents in $s$. For example, if  $\{${\tt{\small (pre\_stack\_holding\_v1),(pre\_stack\_clear\_v2)}}$\} \subset s$ then it must be checked that positive literals {\tt{\small (holding blockB)}} and {\tt{\small (clear blockA)}} hold in $s$. Otherwise, a different set of precondition literals will be checked. The same applies to the conditional effects, generating the corresponding literals according to the values of the model fluents of $s$.

@@

\begin{figure}[hbt!]
\begin{center}
\begin{scriptsize}
\begin{verbatim}
(:action apply_stack
  :parameters (?o1 - object ?o2 - object)
  :precondition (and ) 
  :effect (and (when (and (pre_stack_on_v1_v1) (eff_stack_on_v1_v1)) (not (on ?o1 ?o1)))
               (when (and (pre_stack_on_v1_v2) (eff_stack_on_v1_v2)) (not (on ?o1 ?o2)))
               (when (and (pre_stack_on_v2_v1) (eff_stack_on_v2_v1)) (not (on ?o2 ?o1)))
               (when (and (pre_stack_on_v2_v2) (eff_stack_on_v2_v2)) (not (on ?o2 ?o2)))
               (when (and (pre_stack_ontable_v1) (eff_stack_ontable_v1)) (not (ontable ?o1)))
               (when (and (pre_stack_ontable_v2) (eff_stack_ontable_v2)) (not (ontable ?o2)))
               (when (and (pre_stack_clear_v1) (eff_stack_clear_v1)) (not (clear ?o1)))
               (when (and (pre_stack_clear_v2) (eff_stack_clear_v2)) (not (clear ?o2)))
               (when (and (pre_stack_holding_v1) (eff_stack_holding_v1)) (not (holding ?o1)))
               (when (and (pre_stack_holding_v2) (eff_stack_holding_v2)) (not (holding ?o2)))
               (when (and (pre_stack_handempty) (eff_stack_handempty)) (not (handempty)))
               (when (and (not (pre_stack_on_v1_v1)) (eff_stack_on_v1_v1)) (on ?o1 ?o1))
               (when (and (not (pre_stack_on_v1_v2)) (eff_stack_on_v1_v2)) (on ?o1 ?o2))
               (when (and (not (pre_stack_on_v2_v1)) (eff_stack_on_v2_v1)) (on ?o2 ?o1))
               (when (and (not (pre_stack_on_v2_v2)) (eff_stack_on_v2_v2)) (on ?o2 ?o2))
               (when (and (not (pre_stack_ontable_v1)) (eff_stack_ontable_v1)) (ontable ?o1))
               (when (and (not (pre_stack_ontable_v2)) (eff_stack_ontable_v2)) (ontable ?o2))
               (when (and (not (pre_stack_clear_v1)) (eff_stack_clear_v1)) (clear ?o1))
               (when (and (not (pre_stack_clear_v2)) (eff_stack_clear_v2)) (clear ?o2))
               (when (and (not (pre_stack_holding_v1)) (eff_stack_holding_v1)) (holding ?o1))
               (when (and (not (pre_stack_holding_v2)) (eff_stack_holding_v2)) (holding ?o2))
               (when (and (not (pre_stack_handempty)) (eff_stack_handempty)) (handempty))
               (when (and (pre_stack_on_v1_v1) (not (on ?o1 ?o1))) (invalid))
               (when (and (pre_stack_on_v1_v2) (not (on ?o1 ?o2))) (invalid))
               (when (and (pre_stack_on_v2_v1) (not (on ?o2 ?o1))) (invalid))
               (when (and (pre_stack_on_v2_v2) (not (on ?o2 ?o2))) (invalid))
               (when (and (pre_stack_ontable_v1) (not (ontable ?o1))) (invalid))
               (when (and (pre_stack_ontable_v2) (not (ontable ?o2))) (invalid))
               (when (and (pre_stack_clear_v1) (not (clear ?o1))) (invalid))
               (when (and (pre_stack_clear_v2) (not (clear ?o2))) (invalid))
               (when (and (pre_stack_holding_v1) (not (holding ?o1))) (invalid))
               (when (and (pre_stack_holding_v2) (not (holding ?o2))) (invalid))
               (when (and (pre_stack_handempty) (not (handempty))) (invalid))
               (when (modeProg) (not (modeProg)))))
\end{verbatim}
\end{scriptsize}
 \caption{\small PDDL action for applying an already programmed model for $stack$ (implications are coded as disjunctions).}
\label{fig:compilation}
\end{center}
\end{figure}

@@

\item Actions for {\em validating} the partially observed state $s_j\in\mathcal{O}$, {\tt\small $1\leq j< m$}. These actions are also part of the postfix of the solution plan $\pi_\Lambda$ and they are aimed at checking that the observation $\mathcal{O}$ follows after the execution of the apply actions.

@@

There will be a validate action in $\pi_\Lambda$ for every observed state in $\mathcal{O}$. The position of the validate actions in $\pi_\Lambda$ will be determined by the planner by checking that the state resulting after the execution of an apply action comprises the observed state $s_j\in\mathcal{O}$.

@@

In some contexts, it is reasonable to assume that some parts of the action model are known and so there is no need to learn the entire model from scratch \cite{ZhuoNK13}. In our compilation approach, when an action model $\xi$ is partially specified, the known preconditions and effects are encoded as fluents $pre_{p,\xi}$ and $eff_{p,\xi}$ set to true in the initial state $I_{\Lambda}$. In this case, the corresponding insert actions, $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertEff_{p,\xi}}$, become unnecessary and are removed from $A_{\Lambda}$, thereby making the classical planning task $P_{\Lambda}$ easier to be solved.

So far we explained the compilation for learning from a single input trace. However, the compilation is extensible to the more general case $\Lambda=\tup{\mathcal{M},\mathcal{T}}$, where $\mathcal{T}=\{\mathcal{O}_1,\ldots,\mathcal{O}_k\}$ is a set of $k$ observations. Taking this into account, a small modification is required in our compilation approach. In particular, the actions in $P_{\Lambda}$ for {\em validating} the last state $s_{m,t}^o\in \mathcal{O}_t$, {\tt\small $1\leq t\leq k$} of an observation $\mathcal{O}_t$ reset the current state. These actions are now redefined as:

\begin{small}
	\begin{align*}
	\hspace*{7pt}\pre(\mathsf{validate_{j}})=&s_{m,t}^o\cup\{test_{j-1}\}\cup \{\neg mode_{prog}\},\\
	\cond(\mathsf{validate_{j}})=&\{\emptyset\}\rhd\{\neg test_{j-1},test_j\} \cup \\
	&\{\neg f\}_{\forall f\in s_{m,t}^o, f \notin s_{0,t+1}^o}\cup \{f\}_{\forall f\in s_{0,t+1}^o, f \notin s_{m,t}^o}.
	\end{align*}
\end{small}

Finally, we will detail the composition of a solution plan $\pi_\Lambda$ to a planning task $P_\Lambda$ and the mechanism to extract the action models of $\mathcal{M}'$ from $\pi_\Lambda$. The plan of Figure~\ref{fig:plan-lplan} shows a solution to the task $P_{\Lambda}$ that encodes a learning task $\Lambda=\tup{\mathcal{M},\mathcal{O}}$ for obtaining the action models of the {\em blocksworld} domain, where the models for {\tt\small pickup}, {\tt\small putdown} and {\tt\small unstack} are already specified in $\mathcal{M}$. Therefore, the plan shows the insert actions and validate action for the action model {\tt\small stack}. Plan steps $00-01$ insert the preconditions of the {\tt\small stack} model, steps $02-06$ insert the action model effects, and steps $07-11$ form the plan postfix that applies the action models (only the {\tt\small stack} model is learned) and validates the result in the input observation.

\begin{figure}[hbt!]
	{\footnotesize\tt
		{\bf 00} : (insert\_pre\_stack\_holding\_v1) \\
		01 : (insert\_pre\_stack\_clear\_v2)\\
		{\bf 02} : (insert\_eff\_stack\_clear\_v1)\\
		03 : (insert\_eff\_stack\_clear\_v2)\\
		04 : (insert\_eff\_stack\_handempty)\\
		05 : (insert\_eff\_stack\_holding\_v1)\\
		06 : (insert\_eff\_stack\_on\_v1\_v2)\\
		{\bf 07} : (apply\_unstack blockB blockA i1 i2)\\
		08 : (apply\_putdown blockB i2 i3)\\
		09 : (apply\_pickup blockA i3 i4)\\
		10 : (apply\_stack blockA blockB i4 i5)\\
		{\bf 11} : (validate\_1)
	}
	\caption{\small Plan for programming the $stack$ action model and for validating the programmed $stack$ action model with previously specified action models for $pickup$, $putdown$ and $unstack$.}
	\label{fig:plan-lplan}
\end{figure}

Given a solution plan $\pi_\Lambda$ that solves $P_{\Lambda}$, the set of action models $\mathcal{M}'$ that solves $\Lambda=\tup{\mathcal{M},\mathcal{O}}$ learning task are computed in linear time and space. In order to do so, $\pi_\Lambda$ is executed in the initial state $I_{\Lambda}$ and the action model $\mathcal{M}'$ will be given by the fluents $pre_{p,\xi}$, and $eff_{p,\xi}$ that are set to true in the last state reached by $\pi_\Lambda$, $s_g=\theta(I_\Lambda,\pi_\Lambda)$. For each $\xi \in \mathcal{M'}$, we build the sets of preconditions, positive effects and negative effects as follows:
